\bf Model Visualization Techniques for a Social Network Model

### 2017-03-29 paper.Rnw
add ref to dyn net viz

The concept of model visualization was developed to complement traditional model diagnostic tools. Typically, numerical summaries such as $R^2$ are used to assess model fit, and visualizations, like residual plots, are restricted to the data. Every good data analysis includes both numerical and visual summaries of the data, so why should model description be restricted to numerical summaries? \citeauthor{modelvis} outline the three different parts of a model: the model family, the model form, and the fitted model. The latter is primarily what one thinks of first when considering a model in a data analysis, where a model is fit to the data, and parameter estimates reported. In the context of SAOMs, the fitted model contains the estimated rate parameters and the estimated objective function parameters. The model form describes the the model before the fitting process and how variables describe the data in the model. In SAOMs, the model form includes description of the rate and objective functions and the variables therein that describe how the network evolves over time. Finally, the model family is the most broad description of the model. This is the type of model that you wish to fit to the data, and is chosen based on the problem and knowledge at had. For example, we chose to use a SAOM to model network data over an exponential random graph model (ERGM) because we believe that actor-level variables effect network dynamics. These three parts of a "model" are should be visualized according to the three principals of model visualization: display the model in the data space, look at a collection of models, and explore the process of fitting the model, not just the end result.

@@

A key component of each step of the SIENA method of moments algorithm is the ``microstep." This step simulates from the model in its current state, $x(t_m)$ with current parameter values $\theta_{0}$, to the next state, $x(t_{m+1})$, or as close as it can get to that state. This simulation process follows the steps of the continuous-time Markov chain. First, an ``ego node'' is selected to make a change, and then it randomly makes one change in its ties according to the probability, $p_{ij}$, determined by its objective function. The options for change are (1) removing a current tie, (2) adding a new tie, or (3) making no change at all. Saving all of these steps is not efficient if you're only interested in parameter values, but they can be saved and extracted using options in \texttt{RSiena}, which is exactly what we did to create our visualizations. Between two network observations $x(t_m)$ and $x(t_{m+1})$, there can be hundreds or even thousands of microsteps. We wanted to see these in-between steps in order to better understand the behavior of the underlying continuous-time Markov chain.

The first vizualization we present here is an animation of microsteps simulated to emulate the transition steps from wave 1 to wave 2 the subset of the friends data shown in Figure ~\ref{fig:alldataviz}. When each ego node is selected in a microstep, it is emphasized in the animation, then the associated edge either appears or disappears. If there are no changes at a particular microstep, no changes are seen. Some frames of the animations are shown in INSERT STATIC FIGURE OF THE 5-6 DIFFERENT MICROSTEPS AND ONE TRANSITION SET, and the full movie can be viewed at INSERT LINK TO GIF.

>>>>

The concept of model visualization was developed to complement traditional model diagnostic tools. Typically, numerical summaries such as $R^2$ are used to assess model fit, and visualizations, like residual plots, are restricted to the data. Every good data analysis includes both numerical and visual summaries of the data, so why should model description be restricted to numerical summaries? \citeauthor{modelvis} outline the three different parts of a model: the model family, the model form, and the fitted model. The latter is primarily what one thinks of first when considering a model in a data analysis, where a model is fit to data, and parameter estimates are reported. In the context of SAOMs, the fitted model contains the estimated rate parameters and the estimated objective function parameters. The model form describes the the model before the fitting process and how variables describe the data in the model. In SAOMs, the model form includes description of the rate and objective functions and the variables therein that describe how the network evolves over time. Finally, the model family is the most broad description of the model. This is the type of model that you wish to fit to the data, and is chosen based on the problem and knowledge at had. For example, we chose to use a SAOM to model network data over an exponential random graph model (ERGM) because we believe that actor-level variables effect network dynamics. These three parts of a "model" are should be visualized according to the three principals of model visualization: display the model in the data space, look at a collection of models, and explore the process of fitting the model, not just the end result.

@@

A key component of each step of the SIENA method of moments algorithm is the ``microstep." This step simulates from the model in its current state, $x(t_m)$ with current parameter values $\theta_{0}$, to the next state, $x(t_{m+1})$. This microstep process stops when the simulated network has acheived the same number of differences from $x(t_m)$ as $x(t_{m+1})$. This simulation process follows the steps of the continuous-time Markov chain. First, an ``ego node'' is selected to make a change, and then it randomly makes one change in its ties according to the probability, $p_{ij}$, determined by its objective function. The options for change are (1) removing a current tie, (2) adding a new tie, or (3) making no change at all. Saving all of these steps is not efficient if you're only interested in parameter values, but they can be saved and extracted using options in \texttt{RSiena}, which is exactly what we did to create our visualizations. Between two network observations $x(t_m)$ and $x(t_{m+1})$, there can be dozens, hundreds, or even thousands of microsteps. We wanted to see these in-between steps in order to better understand the behavior of the underlying continuous-time Markov chain.

The first vizualization we present here is an animation of microsteps simulated to emulate the transition steps from wave 1 to wave 2 the subset of the friends data shown in Figure ~\ref{fig:alldataviz}. Movies similar to our animation were used to visualize the changes of dynamic networks in \citet{dynnetviz}. When each ego node is selected in a microstep, it is emphasized in the animation, then the associated edge either appears or disappears. If there are no changes at a particular microstep, no changes are seen. Some frames of the animations are shown in INSERT STATIC FIGURE OF THE 5-6 DIFFERENT MICROSTEPS AND ONE TRANSITION SET, and the full movie can be viewed at INSERT LINK TO GIF.
